---
title: "lab 6"
author: "J Brophy"
date: '2019-01-14'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab # 6

Recall from previous labs where we used the `heart` data and the `table` function which builds a contingency table of the counts at each combination of factor levels. The `epi.2by2` function computes summary measures of risk and a chi-squared test for difference in the observed proportions
from count data presented in a 2 by 2 table. With multiple strata the function returns crude
and Mantel-Haenszel adjusted measures of association and chi-squared tests of homogeneity.
```{r}
# from last week
#read the dataset
library(epiR)

heart <- read.csv("heart.csv")

#create binary variables
heart$female <- factor(heart$gender, labels=c("female", "male"), levels= c(1,0))
heart$fstat <- factor(heart$fstat, labels=c("died", "alive"), levels= c("dead", "alive"))
heart$cvd <- factor(heart$cvd, labels=c("yes", "no"), levels=c("yes", "no"))

tabMH <- table(heart$cvd, heart$fstat, heart$female, dnn = c("CVD", "Died", "Female"))
tabMH

table_MH <-  epi.2by2(tabMH, method = "cohort.count")
table_MH

# Calculating the adjusted PAF using the first method (stratification + weighted average)
#percent of cases that are male/female
percent <- table(heart$female, heart$fstat) 
percent

#females are first, so we get the % of females cases
#females with outcome over all cases
percent_female <- (percent[[1,1]])/(percent[[1,1]]+percent[[2,1]]) 
percent_female

#males are second, so we get the % of males cases
#males with outcome over all cases
percent_male <- (percent[[2,1]])/(percent[[1,1]]+percent[[2,1]]) 
percent_male

table_MH$massoc$PAFRisk.strata

#save the PAF variables to "objects"
female_PAF_RR <- table_MH$massoc$PAFRisk.strata[1,1]
male_PAF_RR <- table_MH$massoc$PAFRisk.strata[2,1]

PAF_RR = percent_female*female_PAF_RR + percent_male*male_PAF_RR
PAF_RR

```

Again from last week

```{r adjusted IRR}
lenfol <- tapply(heart$lenfol, list(heart$cvd, heart$female), sum)
lenfol
events <- tapply(heart$fstat=="died", list(heart$cvd, heart$female), sum)
events

IRR_Female <- array(c(83, 21, 132067, 33425, 87, 24, 191427, 84299),
                   dim=c(2,2,2),
                   dimnames=list(CVD = c("Prior CVD", "No Prior CVD"),
                                 Outcome = c("Events", "Length of FU"),
                                 Covariate = c("Female", "Male")))
IRR_Female
IRR_Female_MH <- epi.2by2(IRR_Female, method="cohort.time")
IRR_Female_MH$massoc$IRR.strata
IRR_Female_MH

IRR_Female_MH$massoc$PAFRate.strata.wald
female_PAF_IRR <- IRR_Female_MH$massoc$PAFRate.strata.wald[1,1]
male_PAF_IRR <- IRR_Female_MH$massoc$PAFRate.strata.wald[2,1]

PAF_IRR = percent_female*female_PAF_IRR[[1]] + percent_male*male_PAF_IRR[[1]]
PAF_IRR
```

```{r}

# method 2 

percent_exp <- table(heart$cvd, heart$fstat)
percent_exp

percent_exp_cases <- percent_exp[[1,1]]/(percent_exp[[1,1]] + percent_exp[[2,1]])
percent_exp_cases
```

Get the MH pooled RR
Recall that table_MH is the table of: exposed: prior CVD, outcome: dead/alive, covariate: gender 

```{r}
table_MH
```

We can save the MH-adjusted RR and use it in Meittenan’s equation    
$$ PAF = \dfrac{p_{ce} * (RR -1)}{RR}$$

```{r}
#save the pooled RR as an object
table_MH <- epi.2by2(tabMH, method = "cohort.count") 
RR_MH <- table_MH$massoc$RR.mh[[1]]
RR_MH

#Now calculate Meittenan's equation
PAF2 <- ((percent_exp_cases)*(RR_MH-1))/RR_MH 
PAF2
```

## Bias adjustment

This lab will explore bias adjustments. The primary packages to be used are `episensr` and `epiR`. `episensr`
contains functions for ```basic sensitivity analysis of the observed relative risks adjusting
for unmeasured confounding and misclassification of the
exposure/outcome, or both. It follows the bias analysis methods and
examples from the book by Lash T.L, Fox M.P, and Fink A.K.
``Applying Quantitative Bias Analysis to Epidemiologic Data''.``` 

```{r}
library(episensr)

lowbirth <- read.csv("lowbirth.csv")
head(lowbirth)

# making factors in the right order

lowbirth$smoke_fac <- factor(lowbirth$smoke, labels=c("Smoker", "Non-smoker"), levels=c(1,0))
lowbirth$low_fac <- factor(lowbirth$low, labels=c("Low", "Normal"), levels=c(1,0))

out_table <- selection(case=lowbirth$low_fac, exposed=lowbirth$smoke_fac, bias_parms = c(.9, .8, .4, .5))
out_table
# can also see what the modified (corrected) table looks like
str(out_table)
out_table$corr.data
```
## Exercise 1

A team conducted a case control study on CVD and cheeseburger consumption. 
```{r}
cvd <- matrix(c(845, 1765, 4478, 6942), nrow=2, ncol=2, byrow=TRUE)
rownames(cvd) <- c("Cases", "Controls")
colnames(cvd) <- c("Exposed","Unexposed")

```
You find out that the probabilities of selecting cases and controls into the study are: 
 30% among cases
70% among controls
How much bias  was there in the original study?

```{r}
selection(cvd, bias_parms=c(.3, .3, .7, .7))

```


OR observed == OR bias corrected

Is this a mistake?
No! The observed estimate (OR=0.74) was unbiased (corrected OR=0.74) because selection probabilities were non-differential w.r.t. exposure status, meaning that exposed cases were selected with the same probability as unexposed unexposed cases. 

## Misclassification

Can do the same exercise if exposure or outcome status was misclassified
```{r}
table(lowbirth$low_fac, lowbirth$smoke_fac)

# To get true cell counts from observed cell counts, enter misclassification as: c(Sensitivity in exposed, Sensitivity in unexposed, Specificity in exposed, Specificity in unexposed)

misclassification(exposed=lowbirth$smoke_fac, case=lowbirth$low_fac, bias_parms=c(.9, .9, .8, .8), type="exposure")

```

The observed OR was biased downward or towards the null. This is because the degree of misclassification of exposure is equivalent in the diseased vs non-diseased groups (i.e., non-differential misclassification of exposure).

```{r}
mis <- misclassification(exposed=lowbirth$smoke_fac, case=lowbirth$low_fac, bias_parms=c(.9, .9, .8, .8), type="exposure")
mis$corr.data
```


## Uncontrolled confounding

In this case-control study on LBW relative to smoking, suppose information maternal education is missing
How much bias is introduced by failing to include maternal education in the analysis?
Assumptions:
OR relating ME to LBW is 1.5 (from the literature)
Prevalence of higher education among smokers=0.3
Prevalence of higher education among non-smokers = 0.5

```{r}
confounders(exposed=lowbirth$smoke_fac, case=lowbirth$low_fac, type=c("OR"), bias_parms = c(1.5, 0.3, 0.5))

```
The observed estimate (OR=2.88) was biased downward (corrected OR=3.14). Therefore, omitting to include maternal education resulted in an 8% bias towards the null.       

Residual confounding may occur if: 
Unmeasured,
Inaccurately measured
Misclassified
Crudely categorized confounders
Eg, Collapsing continuous variables into categories
Sometimes can assess impact of residual confounding by changing categorization of confounder (creating finer groups)

## Exercise 2
Using LBW & smoking dataset, create a categorical variable for age with two and three levels:
	a) two levels: < 22 and ≥22
	b) three levels: < 18, ≥18 to <25; ≥ 25
In both situations, compute the odds ratio standardized to the age distribution in the exposed group.
Is there evidence of residual confounding by age?

```{r}
lowbirth$agecat2 <- factor(cut(lowbirth$age, c(0, 22, Inf), right=FALSE), labels=c("under 22", "22+"))

lowbirth$agecat3 <- factor(cut(lowbirth$age, c(0, 18, 25, Inf), labels=c("under 18", "over 18, under 25","25+"), right=FALSE))

tab_smoke_low_age2 <- with(lowbirth, table(smoke_fac, low_fac, agecat2))
epi_stand_age2 <- epi.2by2(tab_smoke_low_age2, method="case.control")

epi_stand_age2
epi_stand_age2$massoc$OR.strata.wald

```


 
## Adjusting using standardized measures

```{r}
install.packages("epimisc_0.12.tar.gz", repos = NULL, type = "source")
library(epimisc)
std.measure(tab_smoke_low_age2, conflev=0.95, method="internal", measure="risk.ratio")

#3 level variable table
tab_smoke_low_age3 <- with(lowbirth, table(smoke_fac, low_fac, agecat3))

epi_stand_age3 <- epi.2by2(tab_smoke_low_age3, method="case.control")
epi_stand_age3
# to get individual strata
epi_stand_age3$massoc$OR.strata.wald

#to get standardized measure. Only available for risk ratio and risk difference currently.
library(epimisc)
?std.measure
std.measure(tab_smoke_low_age3, conflev=0.95, method="internal", measure="risk.ratio")

```


```{r}
#3 level variable table
tab_smoke_low_age3 <- with(lowbirth, table(smoke_fac, low_fac, agecat3))

epi_stand_age3 <- epi.2by2(tab_smoke_low_age3, method="case.control")
epi_stand_age3
# to get individual strata
epi_stand_age3$massoc$OR.strata.wald

#to get standardized measure. Only available for risk ratio and risk difference currently.
library(epimisc)
std.measure(tab_smoke_low_age3, conflev=0.95, method="internal", measure="risk.ratio")

```

