---
title: "Lab2 Binary logistic regression"
author: " "
date: ' '
output: html_document
---

## Context

```{r setup, include=FALSE}
options(width = 200)
knitr::opts_chunk$set(echo =TRUE, comment = NA, message=FALSE, warning = FALSE, error=FALSE)
```

A cross sectional study was conducted to investigate the association between different activity levels and occurrence of myopia among school children of age between 5-9. 618 children were recruited and following variables were measured:  

variable name | description
---------------|------------
age | Age (Years)
gender | Gender (1: Male, 2: Female)
spheq | Spherical Equivalent Refraction (diopter)
al | Axial length (mm)
acd | Anterior Chamber Depth (mm)
lt | Lens Thickness (mm)
vcd | Vitreous Chamber Depth (mm)
sporthr | How many hours per week outside of school the child spent engaging in sports/outdoor activities (Hours per week)   
readhr | How many hours per week outside of school the child spent reading for pleasure (Hours per week)     
comphr | How many hours per week outside of school the child spent playing video/computer games or working on the computer (Hours per week)     
studyhr | How many hours per week outside of school the child spent reading or studying for school assignments (Hours per week)    
tvhr | How many hours per week outside of school the child spent watching television (Hours per week)    
diopterhr | Composite of near-work activities (Hours per week)     
mommy | Was the subject's mother myopic? (1: No, 2: Yes)    
dadmy | Was the subject's father myopic? (1: No, 2: Yes)  
 
The data is available from the 'aplore' package. 

# Research Question  
**To estimate the extend to which hours spend outside school engagin in sports/outdoor activity is associated with risk of myopia among children below 10 years of age**

### Loading necessary libraries   
```{r}
library(Hmisc)
library(tidyverse)
library(tableone)
library(stargazer)
library(aplore3)
```

### Importing the data
```{r}
data(myopia)
```

```{r}
skimr::skim(myopia)
```
There are no missing values in the dataset. There are 4 categorical variables and 12 continuous variables. There are some large values in sport/outdoor activities hours (40!).

## Descriptive statistics
### Continuous variables
```{r}
layout(matrix(c(1:12), nrow=2, ncol=3, byrow = F))
for(i in c("age","spheq","al","acd","lt","vcd","sporthr","readhr","comphr","studyhr","tvhr","diopterhr")){
  boxplot(myopia[[i]], main=paste("Box plot of", i))
  hist(myopia[[i]], main=paste("Histogram of", i), xlab=paste(i))
}
```

+ Most of the variables have outliers 

### Categorical variables
```{r}
tab1 <- CreateTableOne(data=myopia, 
                       vars=c("myopic","gender","mommy","dadmy"), includeNA = TRUE)
print(tab1, showAllLevels=TRUE)
```

+ Only 81 (13\%) participants have myopia. 

## Bivariate relationships
### Correlations
Because most of the continuous variables are not normally distributed we will use spearman's correlation. 
```{r}
corr <- myopia %>% 
  select(age,spheq,al,acd,lt,vcd,sporthr,readhr,comphr,studyhr,tvhr,diopterhr) %>%
  cor(method="spearman") %>%
  round(digits = 2)
corr
```

**BEING FANCY**
```{r}
corrplot::corrplot(corr, method="color")
```

+ vcd is highly correlated with al
+ Diopterhr variable seems to have moderate level correlations with some other activity variables


## Contegency Tables
```{r}
descr::CrossTable(myopia$gender, myopia$mommy, prop.t=F, prop.chisq = F)
descr::CrossTable(myopia$gender, myopia$dadmy, prop.t=F, prop.chisq=F)
descr::CrossTable(myopia$mommy, myopia$dadmy, prop.t=F, prop.chisq=F)
```

**You can also easly see these distributions quickly by plotting**
```{r}
par(mfrow=c(1,3))
plot(table(myopia$gender, myopia$mommy))
plot(table(myopia$gender, myopia$dadmy))
plot(table(myopia$dadmy, myopia$mommy))
```

## DAG

Assuming that there are no unmeasured confounding between myopic and other variables in the data. We can construct a DAG to see what is the minimum adjustment set we need. 
```{r}
library(ggdag)

coords = list(
        x = c(mommy=2, dadmy=2, sporthr=2, readhr=2, comphr=2, studyhr=2, tvhr=2, diopterhr=2, myopic=4,
              age=1, gender=1, spheq=3, al=3, acd=3, lt=3, vcd=3),
        y = c(mommy=-4, dadmy=-3, readhr=-2, comphr=-1, sporthr=0, studyhr=1, tvhr=2, diopterhr=3, myopic=0,
              age=-1, gender=1, spheq=-4, al=-3, acd=-2, lt=2, vcd=3)
)

dag1 <- dagify(myopic ~ age+gender+mommy+dadmy+sporthr+readhr+comphr+studyhr+tvhr+diopterhr+spheq+al+acd+lt+vcd,
               mommy ~~ sporthr ,
               dadmy ~~ sporthr,
               sporthr ~~ readhr,
               sporthr ~~ comphr,
               sporthr ~~ studyhr,
               sporthr ~~ tvhr,
               sporthr ~~ diopterhr,
               sporthr ~ age+gender,
               readhr ~ age+gender,
               comphr ~ age+gender,
               studyhr ~ age+gender,
               tvhr ~ age+gender,
               diopterhr ~ age+gender,
               spheq ~ age+gender+sporthr,
               al ~ age+gender+sporthr,
               acd ~ age+gender+sporthr,
               lt ~ age+gender+sporthr,
               vcd ~ age+gender+sporthr,
              exposure ="sporthr",
              outcome = "myopic", coords = coords)

ggdag(dag1, node_size = 20)
ggdag_adjustment_set(dag1)
```

Because we are interested in the total effect of sporthr, we will not adjust for the descendants of sporthr. 

## Table 1
We shall start with looking at the distribution of all variables across the outcome.
```{r}
tab2 <- CreateTableOne(vars=c("age","gender","sporthr","readhr","comphr","studyhr","tvhr", "diopterhr", "mommy","dadmy"), strata = "myopic", data=myopia, includeNA=TRUE, test=FALSE)
print(tab2, showAllLevels=TRUE)
```

+ Myopic kids had more average hours in sport/outdoor activity
+ There are slightly more Female kids among myopic group
+ Myopic kids had high proportion of participants with family history of myopia

## Univariate logistic regression
```{r}
uniregs <- list()
uniregs[[1]] <- glm(myopic ~ age, data=myopia, family=binomial(link="logit"))
uniregs[[2]] <- glm(myopic ~ gender, data=myopia, family=binomial(link="logit"))
uniregs[[3]] <- glm(myopic ~ sporthr, data=myopia, family=binomial(link="logit"))
uniregs[[4]] <- glm(myopic ~ readhr, data=myopia, family=binomial(link="logit"))
uniregs[[5]] <- glm(myopic ~ comphr, data=myopia, family=binomial(link="logit"))
uniregs[[6]] <- glm(myopic ~ studyhr, data=myopia, family=binomial(link="logit"))
uniregs[[7]] <- glm(myopic ~ tvhr, data=myopia, family=binomial(link="logit"))
uniregs[[8]] <- glm(myopic ~ diopterhr, data=myopia, family=binomial(link="logit"))
uniregs[[9]] <- glm(myopic ~ mommy, data=myopia, family=binomial(link="logit"))
uniregs[[10]] <- glm(myopic ~ dadmy, data=myopia, family=binomial(link="logit"))

stargazer(uniregs[1:5], type="text", report="vcs", ci=TRUE, omit.table.layout = "n", digits=2)
stargazer(uniregs[6:10], type="text", report="vcs", ci=TRUE, omit.table.layout = "n", digits=2)
```

+ age, gender(female), readhr, comphr, diopterhr, mommy, dadmy show positive association with myopic state
+ sporthr show negative association
+ interestingly, studyhr and tvhr show a negative association!

## Multivariable logistic regression
```{r}
glm1 <- glm(myopic ~ age + gender + mommy + dadmy + 
              sporthr + readhr + comphr + studyhr + tvhr + diopterhr , 
            data=myopia, family=binomial(link="logit"))
stargazer(glm1, type="text", report="vcs", ci=TRUE, omit.table.layout = "n", digits=2)
```

**The diopterhr variable do not have a coefficient. Probably due to high collinearity R took it out**  
**We need to report the effect measures in Odds ratio scale**
```{r}
glm1 <- glm(myopic ~ age + gender + mommy + dadmy + 
              sporthr + readhr + comphr + studyhr + tvhr , 
            data=myopia, family=binomial(link="logit"))
ors <- exp(coef(glm1))
cis <- exp(confint.default(glm1))

stargazer(glm1, type="text", report="vcs", coef=list(ors), ci.custom = list(cis), omit.table.layout = "n", digits=2)
```

## Interpretation 
Each hour increase in hours spend outside school, engaging in sports/outdoor activities, per week decreased the odds of myopia by a factor of 0.96 (OR = 0.06, 95\%CI = 0.93, 0.99). 

# Contrasts

+ What is the relative odds of myopia for a male child with both parents having with myopia, compared to a female child with no family history of myopia
  + Create the two profiles first
  + Contrast the profiles

```{r}
a = list(age=median(myopia$age), 
         gender="Male",
         mommy="Yes", 
         dadmy="Yes",
         sporthr=median(myopia$sporthr),
         readhr=median(myopia$readhr),
         comphr=median(myopia$comphr),
         studyhr=median(myopia$studyhr),
         tvhr=median(myopia$tvhr))

b = list(age=median(myopia$age),
         gender="Female",
         mommy="No",
         dadmy="No",
         sporthr=median(myopia$sporthr),
         readhr=median(myopia$readhr),
         comphr=median(myopia$comphr),
         studyhr=median(myopia$studyhr),
         tvhr=median(myopia$tvhr))

c1 <- contrast::contrast(glm1,a=a,b=b)
c1
```

**You can equivalently create contrast setting other variables to zero**
```{r, eval=FALSE}
a = list(age=0,sporthr=0,readhr=0,comphr=0,studyhr=0,tvhr=0, gender="Male", mommy="Yes", dadmy="Yes")
b = list(age=0,sporthr=0,readhr=0,comphr=0,studyhr=0,tvhr=0, gender="Female", mommy="No", dadmy="No")
```

**Contrast in odds ration scale**
```{r}
exp(cbind(OR=c1$Contrast, lower=c1$Lower, upper=c1$Upper)) %>% round(digits = 2)
```

+ What is the effect each 5 hour increase in sports time?
  + set all other variables to its reference category
```{r}
a1 = list(age=0,sporthr=5,readhr=0,comphr=0,studyhr=0,tvhr=0, gender="Male", mommy="No", dadmy="No")
b1 = list(age=0,sporthr=0,readhr=0,comphr=0,studyhr=0,tvhr=0, gender="Male", mommy="No", dadmy="No")

c2 <- contrast::contrast(glm1,a=a1,b=b1)
exp(cbind(OR=c2$Contrast, lower=c2$Lower, upper=c2$Upper)) %>% round(digits = 2)
```